---
title: "Simultaneous Directional Inference"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Supplementary R code for reproducing the examples in the paper ``Simultaneous Directional Inference''. 

## Set up

Install the latest version of the [`nplus `](https://github.com/aldosolari/nplus)  R package:

```{r, results='hide', message=F, warning=F, error=F, comment=NA}
devtools::install_github("aldosolari/nplus")
```

## Introduction

We consider the problem of inference on the signs of $n>1$ parameters. 
Let $\theta = (\theta_1,\ldots,\theta_n)$ be a vector of $n$ unknown real-valued parameters. For any $I\subseteq \{1,\ldots,n\}$, let $n^+(I)$ and  $n^-(I)$ be the number of parameters $\theta_i$, $i \in I$ with positive value and negative values, respectively:
$$
n^+(I) = |\{ i \in I : \theta_i > 0 \}|, \quad n^-(I) = |\{ i \in I : \theta_i < 0 \}|.
$$
Simultaneous inference on $n^+(I)$ and $n^-(I)$ can be obtained by selecting one hypothesis within each pair of one-sided hypotheses
$$
 H_i^-: \theta_i \leq 0, \qquad
 H_i^+: \theta_i \geq 0.
$$
where we have right-sided $p$-values $p_i$ for $H_i^-$ and left-sided $p$-values $q_i=1 - p_i$ for $H_i^+$. We select $H_i^-$ if $p_i \leq 1/2$ and $H_i^+$ if $p_i > 1/2$. We denote by $S$ the vector of signs that we condition on: $$S = (sign(p_1-1/2), \ldots, sign(p_n-1/2)), \quad
 sign(p_i-1/2)  =
\begin{cases}
-1 & \text{if } p_i \leq 1/2, \\
1 & \text{if } p_i > 1/2.
\end{cases}
$$
The *directional closed testing procedure* provides $(1-\alpha)$ confidence bounds $l^+(I)$ and $u^+(I)$ for $n^+(I)$ and confidence bounds $l^-(I)$ and $u^-(I)$ for $n^-(I)$ hat are simultaneous
for all $I$:
$$
\mathrm{P}_{\theta}\Big(
l_{\alpha}^{+}(I)
\leq 
n^{+}(I) 
\leq 
u_{\alpha}^{+}(I),\,\,
l_{\alpha}^{-}(I)
\leq 
n^{-}(I) 
\leq 
u_{\alpha}^{-}(I),
 \mathrm{\,\,for\,\,all\,\,}I \mid  S \Big) \geq 1-\alpha.
$$
If one is interested only in inference on $n^+(I)$, consider the $n$ pairs
$$
 H_i^-: \theta_i \leq 0, \qquad
 K_i: \theta_i > 0.
$$
where $H_i^+$ is modified by removing the equality sign, and thus exactly one of the hypothesis pair $(H_i^-, K_i)$ is true. 
By using this formulation, the *adaptive partitioning procedure* provides
bounds for $n^+(I)$ that are uniformly better than those of directional closed testing:
$$
\mathrm{P}_{\theta}\Big(
l_{\alpha}^{+}(I)
\leq 
n^{+}(I) 
\leq 
u_{\alpha}^{+}(I)
 \mathrm{\,\,for\,\,all\,\,}I \mid S \Big) \geq 1- \alpha.
$$

## An example: PFT therapy for breast cancer

The data analysed in Gail and Simon (1985) consists of the
difference in the proportion of patients that are disease free with the PFT treatment versus
with the FT treatment, in $n = 4$ subgroups defined by age and progesterone receptor levels  ($i = 1$: Age < 50, PR
< 10; $i = 2$: Age ≥ 50, PR < 10; $i = 3$: Age < 50, PR ≥ 10; $i = 4$: Age ≥ 50, PR ≥ 10).

```{r}
d_i = c(.163, -.114, -.047, -.151)
s_i = c(.0788, .0689, .0614, .0547)
p_right_tail = 1-pnorm(d_i/s_i)
```

The following Table provides for each subgroup the difference, standard error, and right-tailed $p$-value.

```{r, echo=FALSE, results='asis'}
tab_df = data.frame(rbind(d_i, s_i, p_right_tail))
colnames(tab_df) <- paste("i =",1:4)
rownames(tab_df) <- c("difference", "standard error", "right-tail p-value")
knitr::kable(tab_df, digits = 3)
```

Based on the right-sided $p$-values, we select the hypotheses $H_1^-$, $H_2^+$, $H_3^+$, $H_4^+$ and we adjust for the selection the corresponding $p$-values:

```{r}
n = length(p_right_tail)
S = rep("+",n)
S[p_right_tail <= 0.5] <- "-" 
hyp_selected <- paste0(1:n,S)
p_selected <- 2*pmin(p_right_tail, 1-p_right_tail)
names(p_selected) <- hyp_selected
p_selected
```

Function to compute the power set of a set:

```{r}
powset <- function(set) { 
  n <- length(set)
  masks <- 2^(1:n-1)
  lapply( 1:2^n-1, function(u) set[ bitwAnd(u, masks) != 0 ] )
}
```

Function to compute the $p$-value of the adaptive Simes local test:

```{r}
p_combi_test <- function(x) { ifelse(min(x)<=0.5, ifelse(length(x) < 3, 1, (sum(x>.5)+1)/(length(x)*0.5) ) * min(sort(x)*length(x)/(1:length(x))),1)  }
```

Computing the $p$-value for each intersection hypothesis by using the adaptive Simes local test:

```{r, message=F, warning=F, error=F, comment=NA}
res = round(unlist(
  lapply(powset(1:n), function(x) p_combi_test(p_selected[x]))
),4)
names(res) <- powset(hyp_selected)
res
```

Closed testing adjustment:
```{r}
res_ct = res
closure = powset(1:n)
for (k in 1:length(closure))
res_ct[[k]] = max(res[sapply(1:length(closure), function(i) all(closure[[k]] %in% closure[[i]]))]) 
res_ct
```


Partitioning hypotheses:
```{r}
Sc = rep("-",n)
Sc[p_right_tail <= 0.5] <- "+" 
hyp_part <- powset(1:n)
for (i in 1:length(hyp_part)){
hyp_part[[i]] <- Sc
hyp_part[[i]][ powset(1:n)[[i]] ] <- S[ powset(1:n)[[i]] ]
}
res_part <- res
names(res_part) <- hyp_part
res_part
```



$95\%$ confidence lower and upper bounds for $n^+$ by using the adaptive partitioning procedure with adaptive Simes local tests:
```{r}
library(nplus)
nplus_bound(p_right_tail, alpha=0.05, method="ASimes")
```
Index set of identified positive parameters (i.e. $i:\theta_i >0$) and nonpositive parameters ($i:\theta_i \leq 0$) with familywise error rate control at level $\alpha$:
```{r}
nplus_fwer(p_right_tail, alpha=0.05, method="ASimes")
```

Gail, M. and Simon, R. (1985). Testing for qualitative interactions between treatment
effects and patient subsets. Biometrics, pages 361–372.

## Enhancing meta-analysis

We consider the  data in Konstantopoulos (2011), where the  effect of a modified school calendar (with more frequent but shorter breaks) on student achievement is examined. The meta-analysis uses studies of $n=56$ schools in 11 districts.  The experimental design is a two-group comparison (modified calendar vs traditional
calendar) which involves computing the standardized mean difference
$X_i \sim N(\theta_i,1)$, where $\theta_i>0$ indicates a positive effects, i.e. a higher level of achievement in the
group following the modified school calendar.

We analyze the data by providing the lower and upper bounds on the number of studies with positive effect in all schools and in the (data-driven) top $k$ schools.

Importing data from the R package `metafor `:

```{r, results='hide', message=F, warning=F, error=F, comment=NA}
require("metafor") || install.packages("metafor")
```

```{r, message=F, warning=F, error=F, comment=NA}
dat <- get(data(dat.konstantopoulos2011))
x <- as.vector(dat$yi / sqrt(dat$vi))
n <- length(x)
p_right_tail <- pnorm(x, lower.tail = FALSE)
```

Adaptive partitioning procedure: $95\%$ confidence lower and upper bounds on the number of studies with positive/nonpositive effect (`lo+ `/`lo- `  and `up+ `/`up- `) and corresponding number of discoveries (`id+ `/`id- `) for different combining functions: Fisher, Simes, adaptive Simes and adaptive LRT:

```{r}
RES = matrix(NA, ncol=7, nrow=4)
RES[,1] = c("Fisher","Simes","ASimes","ALRT")
colnames(RES) <- c("Local test", "lo+","up+","id+", "lo-","up-","id-")
for (i in 1:4){
 res_bound <- nplus_bound(p_right_tail, method=RES[i,1])
 res_id <- nplus_fwer(p_right_tail, method=RES[i,1])
 RES[i,2] <- res_bound$lo
 RES[i,3] <- res_bound$up
 RES[i,4] <- length(res_id$positive)
 RES[i,5] <- n-res_bound$up
 RES[i,6] <- n-res_bound$lo
 RES[i,7] <- length(res_id$nonpositive)
}
RES
```
Adaptive partitioning procedure: $95\%$ confidence lower and upper bounds on the number of studies with positive effect in the (data-driven) top $k$ schools with the Fisher's combining function:

```{r}
ix_best <- sort(pmin(p_right_tail,1-p_right_tail), index.return=TRUE)$ix
bestk <- sapply(1:n, function(i) 
  range(which( nplus_pvalue(p_right_tail, method = "Fisher", ix = ix_best[1:i]) > 0.05)) -1
)
plot(asp=1, type="s", ylim=c(0,n),xlim=c(1,n),
     bestk[2,], xlab="k", ylab="95% CI for the top k schools")
lines(bestk[1,],type="s")
abline(h=0)
abline(a=0,b=1)
abline(v=n)
for  (i in 1:n){
  segments(x0=i,x1=i,y0=bestk[1,i], y1=bestk[2,i])
}
```



Directional closed testing procedure: $95\%$ confidence lower and upper bounds on the number of studies with positive/negative effect (`lo+ `/`lo- `  and `up+ `/`up- `) and corresponding number of discoveries (`id+ `/`id- `) for different combining functions: Fisher and Simes:

```{r, results='hide', message=F, warning=F, error=F, comment=NA}
require("hommel") || install.packages("hommel")
require("sumSome") || install.packages("sumSome")
```

```{r}
S_minus = which(p_right_tail <= 0.5)
S_plus = which(p_right_tail > 0.5)
p_twotail = 2*pmin(p_right_tail, 1-p_right_tail)
RES_CT = matrix(NA, ncol=7, nrow=2)
RES_CT[,1] = c("Fisher","Simes")
colnames(RES_CT) <- c("Local test", "lo+","up+","id+", "lo-","up-","id-")
l_plus_sumsome = sumSome::discoveries(sumPvalsPar(g=p_twotail, S = S_minus, type = "Fisher", independence = TRUE))
l_minus_sumsome = sumSome::discoveries(sumPvalsPar(g=p_twotail, S = S_plus, type = "Fisher", independence = TRUE))
RES_CT[1,2]<-l_plus_sumsome
RES_CT[1,3]<-n-l_minus_sumsome
RES_CT[1,5]<-l_minus_sumsome
RES_CT[1,6]<-n-l_plus_sumsome
ix_plus <- sort(p_right_tail, index.return=TRUE)$ix
RES_CT[1,4] <- ifelse(l_plus_sumsome > 0, sum(sapply(1:l_plus_sumsome, function(i)
  sumPvalsPar(g=p_twotail, S=ix_plus[1:i], type = "Fisher", independence = TRUE)$TD) == 1:l_plus_sumsome),
  0)
ix_minus <- sort(1-p_right_tail, index.return=TRUE)$ix
RES_CT[1,7]<- ifelse(l_minus_sumsome>0, sum(sapply(1:l_minus_sumsome, function(i)
  sumPvalsPar(g=p_twotail, S=ix_minus[1:i], type = "Fisher", independence = TRUE)$TD) == 1:l_minus_sumsome),
  0)
res_hommel <- hommel(p_twotail)
l_plus_hommel = hommel::discoveries(res_hommel, ix = S_minus)
l_minus_hommel = hommel::discoveries(res_hommel, ix = S_plus)
RES_CT[2,2]<-l_plus_hommel
RES_CT[2,3]<-n-l_minus_hommel
RES_CT[2,4]<-sum(res_hommel@adjusted[S_minus] <= 0.05)
RES_CT[2,5]<-l_minus_hommel
RES_CT[2,6]<-n-l_plus_hommel
RES_CT[2,7]<-sum(res_hommel@adjusted[S_plus] <= 0.05)
RES_CT
```

Discoveries by Guo and Romano (2015) procedure with FWER control at 0.05 level:

```{r}
GuoRomano2015 <- function(p_right_tail, alpha=0.05){
  
  p <- p_right_tail
  n <- length(p)
  sp <- sort( c(p,1-p), index.return=TRUE )
  REJ <- sp$x <= alpha / (n - 1:n + 1L + alpha)
  REJ_NONPOS <-  1:n %in% sp$ix[REJ]
  REJ_POS <-  (n+1):(2*n) %in% sp$ix[REJ]
  
  return(list(n_discoveries_positive=sum(REJ_NONPOS), n_discoveries_nonpositive=sum(REJ_POS)))
  
}
GuoRomano2015(p_right_tail)
```

Konstantopoulos, S. (2011). Fixed effects and variance components estimation in three-level
meta-analysis. Research Synthesis Methods, 2(1):61–76.

Guo, W. and Romano, J. P. (2015). On stepwise control of directional errors under independence and some dependence. Journal of Statistical Planning and Inference, 163:21–33.

